---
title: "Fouls in Women's Basketball"
output:
  html_notebook: default
  pdf_document: default
  html_document:
    df_print: paged
---

Jack laying out the hypotheses (from Feb. 24, 2017(!)), and I've inserted brackets for the corresponding code/column:

We predict male coaches will act more aggressively the more males referees there are in a game, and will therefore tend to have more male referees present [3 - mean(numfemalerefs)] when they receive technical fouls [filter %>% nummaletechs > 0] than when they do not receive any [filter %>% nummaletechs == 0]. Similarly, we also predict more female referees to be present [mean(numfemalerefs)] when female head coaches receive technical fouls [filter %>% numfemaletechs > 0] than when they do not receive any [filter %>% numfemaletechs == 0]. Thirdly, we also predict the number of technical fouls attributed to female coaches [sum(numfemaletechs)/(sum(nummaletechs) + sum(numfemaletechs)] to be lower than women coaches' makeup of all coaches in NCAA women's basketball [sum(numfemalecoaches)/(nrow()*2)]. Likewise, we expect men head coaches to receive a greater proportion of all the technical fouls [sum(nummaletechs)/(sum(nummaletechs) + sum(numfemaletechs))] than their makeup of all coaches [1-(sum(numfemalecoaches)/(nrow() + nrow()))].

As an exploratory analysis, I'm also curious what the effects would be if we changed the dependent variable from coach technical fouls to total fouls called on the team [numfemaletotalfouls and nummaletotalfouls].

##Import data
```{r}
library(data.table)
library(epitools)
library(dplyr)
library(lme4)

raw.data <- fread('games_and_coaches_for_stats.csv')
```

##Plot some data
```{r}
hist(raw.data$total_fouls)
hist(raw.data$total_techs)
hist(raw.data$numfemalerefs)
```

##Make data file a bit smaller to work with
```{r}
df <- raw.data %>% select(numfemalerefs, total_fouls, total_techs, numfemalecoaches, 
                          away_total_points, away_total_techs, away_total_fouls,
                          home_total_techs, home_total_points, home_total_fouls,
                          Home_Team, Away_Team, coach_last_name_home, coach_last_name_away,
                          ref_1_last_name, ref_2_last_name, ref_3_last_name) %>%
  #I want a binary variable of whether there was a technical foul and whether
  #there was a female ref
  mutate(techfoulbinary = (ifelse(total_techs > 0, 1, 0)), 
         femalerefbinary = factor(ifelse(numfemalerefs > 0, 1, 0)),
         femalecoachbinary = factor(ifelse(numfemalecoaches > 0, 1, 0)),
         malerefbinary = factor(ifelse(nummalerefs > 0, 1, 0)),
         malecoachbinary = factor(ifelse(nummalecoaches > 0, 1, 0)),)

```


## Descriptives stats (with some inferential stats)
One thing we might be interested in is just a simple measure/question:
  - Does having a single female referee increase the risk of a technical foul?
  
To look at this, we might want to use the 'relative risk ratio' (RR). This is 
calculated by taking the number of events in condition A over all events in A 
divided by the number of events in condition B over all events in B.

Not significant whether there was a female ref present or not on whether there was a tech called

```{r}

tab <- matrix(c(sum(((df$techfoulbinary[df$femalerefbinary == 1])==1), na.rm = TRUE),
                sum(((df$techfoulbinary[df$femalerefbinary == 1])==0), na.rm = TRUE),
                sum(((df$techfoulbinary[df$femalerefbinary == 0])==1), na.rm = TRUE),
                sum(((df$techfoulbinary[df$femalerefbinary == 0])==0), na.rm = TRUE)),byrow=TRUE,nrow=2)
mydf <- epitab(tab,method="riskratio")
mydf <- as.data.frame(mydf[1])
row.names(mydf) <- c("Female Ref Present", "Female Ref Absent")
colnames(mydf) <- c("Games w/ Techs", "p0", "Games w/o Techs", "p1", "riskratio", "lower", "upper", "p.value")
mydf
```

It is significant whether there was a male ref present or not, but the risk ratio is apparently extremely small and therefore the effect is negligible (significant in which direction?)

```{r}

tab <- matrix(c(sum(((df$techfoulbinary[df$malerefbinary == 1])==1), na.rm = TRUE),
                sum(((df$techfoulbinary[df$malerefbinary == 1])==0), na.rm = TRUE),
                sum(((df$techfoulbinary[df$malerefbinary == 0])==1), na.rm = TRUE),
                sum(((df$techfoulbinary[df$malerefbinary == 0])==0), na.rm = TRUE)),byrow=TRUE,nrow=2)
mydf <- epitab(tab,method="riskratio")
mydf <- as.data.frame(mydf[1])
row.names(mydf) <- c("Male Ref Present", "Male Ref Absent")
colnames(mydf) <- c("Games w/ Techs", "p0", "Games w/o Techs", "p1", "riskratio", "lower", "upper", "p.value")
mydf
```


```{r}
tab <- matrix(c(sum(((df$techfoulbinary[df$femalecoachbinary == 1])==1), na.rm = TRUE),
                sum(((df$techfoulbinary[df$femalecoachbinary == 1])==0), na.rm = TRUE),
                sum(((df$techfoulbinary[df$femalecoachbinary == 0])==1), na.rm = TRUE),
                sum(((df$techfoulbinary[df$femalecoachbinary == 0])==0), na.rm = TRUE)),byrow=TRUE,nrow=2)
mydf <- epitab(tab,method="riskratio")
mydf <- as.data.frame(mydf[1])
row.names(mydf) <- c("Female Coach Present", "Female Coach Absent")
colnames(mydf) <- c("Games w/ Techs", "p0", "Games w/o Techs", "p1", "riskratio", "lower", "upper", "p.value")
mydf
```

```{r}
tab <- matrix(c(sum(((df$techfoulbinary[df$malecoachbinary == 1])==1), na.rm = TRUE),
                sum(((df$techfoulbinary[df$malecoachbinary == 1])==0), na.rm = TRUE),
                sum(((df$techfoulbinary[df$malecoachbinary == 0])==1), na.rm = TRUE),
                sum(((df$techfoulbinary[df$malecoachbinary == 0])==0), na.rm = TRUE)),byrow=TRUE,nrow=2)
mydf <- epitab(tab,method="riskratio")
mydf <- as.data.frame(mydf[1])
row.names(mydf) <- c("Male Coach Present", "Male Coach Absent")
colnames(mydf) <- c("Games w/ Techs", "p0", "Games w/o Techs", "p1", "riskratio", "lower", "upper", "p.value")
mydf
```


##Linear models

Asks  if there' s at least one female coach, yes or no, at least one female ref, yes or no, and whether there's a technical foul, yes or no; none seem to matter
```{r}
summary(glm(techfoulbinary ~ femalerefbinary*femalecoachbinary, data = df, family = binomial("logit")))
summary(glm(techfoulbinary ~ malerefbinary*femalecoachbinary, data = df, family = binomial("logit")))
summary(glm(techfoulbinary ~ femalerefbinary*malecoachbinary, data = df, family = binomial("logit")))
summary(glm(techfoulbinary ~ malerefbinary*malecoachbinary, data = df, family = binomial("logit")))

```

##More complex relationship?
The more female coaches, the more likely there is to be a technical foul; but it doesn't matter how many female refs there are, and there's no interaction
```{r}
#logistic model
model1 <- (glm(techfoulbinary ~ numfemalerefs*numfemalecoaches, data = df, family = binomial("logit")))
summary(model1)
#odds ratio
exp(model1$coefficients[[2]])
model1 <- (glm(techfoulbinary ~ nummalerefs*nummfemalecoaches, data = df, family = binomial("logit")))
summary(model1)
#odds ratio
exp(model1$coefficients[[2]])
model1 <- (glm(techfoulbinary ~ numfemalerefs*nummalecoaches, data = df, family = binomial("logit")))
summary(model1)
#odds ratio
exp(model1$coefficients[[2]])
model1 <- (glm(techfoulbinary ~ nummalerefs*nummalecoaches, data = df, family = binomial("logit")))
summary(model1)
#odds ratio
exp(model1$coefficients[[2]])
```


##More linear relationship?
Number of female coaches significantly correlated with number of technical fouls called, no significance for number of female refs or the interaction
```{r}
summary(glm(total_techs ~ numfemalerefs*numfemalecoaches, data = df, family=poisson(link="log")))
summary(glm(total_techs ~ nummalerefs*numfemalecoaches, data = df, family=poisson(link="log")))
summary(glm(total_techs ~ numfemalerefs*nummalecoaches, data = df, family=poisson(link="log")))
summary(glm(total_techs ~ nummalerefs*nummalecoaches, data = df, family=poisson(link="log")))

```


```{r}
summary(glm(total_fouls ~ numfemalerefs*numfemalecoaches, data = df, family=poisson(link="log")))
summary(glm(total_fouls ~ nummalerefs*numfemalecoaches, data = df, family=poisson(link="log")))
summary(glm(total_fouls ~ numfemalerefs*nummalecoaches, data = df, family=poisson(link="log")))
summary(glm(total_fouls ~ nummalerefs*nummalecoaches, data = df, family=poisson(link="log")))
```




#Let's account for some things though.
One thing we might want to account for, though is the team.
Maybe a specific team just has more technical fouls called on
them because their coach is an asshole. The models above treated 
every game as an independent event, but we know that isn't true.
A team (coach) is going to have a behavior pattern that is
correlated across time. 

Here we can account for that by including 'team' as a random 
effect. That is, we will account for the correlation pattern
between each of a team's games.

Let's start with our simplest model and see if the presence
of a single female ref on the court will increase the odds of
any techincal fouls being called.

Again, significant for coach gender but that's it
```{r}
model.glmer <- glmer(total_techs ~ numfemalerefs * numfemalecoaches + 
                       (1|Home_Team) + (1|Away_Team), data = df, family=poisson(link="log"))
summary(model.glmer)
```

Okay, now let's see if the number of female refs increase the odds of any technical
fouls being called

No (not significantly).
```{r}
model.glmer2 <- glmer(techfoulbinary ~ numfemalerefs + (1|teamcount), data = df, family = binomial("probit"))
summary(model.glmer2)
```

Now let's see if the number of female refs on the court will affect
the number of technical fouls being called.

No, not significantly when accounting for correlation aomng teams.
```{r}
model.lmer <- lmerTest::lmer(total_techs ~ numfemalerefs + (1|teamcount), data = df)
summary(model.lmer)
```

Maybe this is influenced by the total number of fouls (ie. a rough
total 'aggression' metric). So let's include a random slope for total
number of fouls. Essentially we are going to let the influence of the 
number of female refs on the court on technical fouls vary as a function
of the total number of fouls in each game, and account for the correlation
among teams.

The effect is still small, and not much 'more' significant.
```{r}
model.lmer2 <- lmerTest::lmer(total_techs ~ numfemalerefs + (1+total_fouls|teamcount), data = df)
summary(model.lmer2)
```

However, since the total number of technical fouls is so low,
and obviously not normally distributed (see the above histograms),
we might also want to use a Poisson regression. Poisson regressions
are good for modeling rare events in large datasets.

So, yeah, not much better...
```{r}
model.lmer2.p <- glmer((total_techs) ~ numfemalerefs + (1+total_fouls|teamcount), data = df, family=poisson(link="log"))
summary(model.lmer2.p)
```

Final try. Let's get crazy and fit all the previous models using
a Poisson distribution (this is really the first dataset I ahve ever 
had a good justification for using this distribution!)
```{r}
model.glmer.p <- glmer(techfoulbinary ~ femalerefbinary + (1|teamcount), data = df, family=poisson(link="log"))
summary(model.glmer.p)

model.glmer2.p <- glmer(techfoulbinary ~ numfemalerefs + (1|teamcount), data = df, family=poisson(link="log"))
summary(model.glmer2.p)

model.lmer.p <- glmer(total_techs ~ numfemalerefs + (1|teamcount), data = df, family=poisson(link="log"))
summary(model.lmer.p)

linear.p <- glm(total_techs ~ numfemalerefs, data = df, family=poisson(link="log"))
summary(linear.p)

logistic1.p <- glm(techfoulbinary ~ numfemalerefs, data = df, family=poisson(link="log"))
summary(logistic1.p)

logistic2.p <- glm(techfoulbinary ~ femalerefbinary, data = df, family=poisson(link="log"))
summary(logistic2.p)
```

So these last models leave a resounding, small, but probably 'there' effect.
